---
title: "Logistic Linear Mixed Model (Simplified for Binder)"
author:
  name: Christof Seiler
  affiliation: Department of Statistics, Stanford University
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Goal

Reanalysis of mass cytometry data from @aghaeepour2017immune using the Logistic Linear Mixed Model.

# Prerequisites

Parse input parameters.

```{r parse_input_parameters}
ncells = Inf
zenodo_url = "https://zenodo.org/record/2597291/files/"
prefit = paste0("cytoeffect_llmm_ncells_",ncells,".Rdata")
prefit
prefit_reduced = paste0("cytoeffect_llmm_ncells_",ncells,"_reduced.Rdata")
prefit_reduced
```

Load packages.

```{r load_packages}
library("cytoeffect")
library("tidyverse")
library("magrittr")
library("ggthemes")
library("cowplot")
theme_set(theme_few())
```

# Load Data

Download preprocessed data from Zenodo.

```{r download_zenodo}
rdata_filenames = c(prefit, prefit_reduced)
for(filename in rdata_filenames) {
  if(!file.exists(filename)) {
    download.file(url = paste0(zenodo_url, filename), 
                  destfile = filename,
                  mode = "wb")
  }
}
```

# HMC Diagnostics

Traceplot of posterior samples.

```{r post_sampling, fig.wide=TRUE}
load(file = prefit)
rstan::traceplot(obj$fit_mcmc, inc_warmup = TRUE)
rstan::traceplot(obj$fit_mcmc, inc_warmup = FALSE)
```

Some more MCMC diagnostics. According to empirically findings, Rhat > 1.1 is usually indicative of problems in the fit.

```{r mcmc_diagnostics}
pars = c("beta","sigma_donor","L_donor","z_donor")
tb = rstan::summary(obj$fit_mcmc, 
                    pars = pars)$summary %>% 
  as.tibble(rownames = "pars", .before = 1) %>% 
  select(pars, n_eff, Rhat)
tb %<>% na.omit() # Stan fills upper triangle with zeros
tb %>% arrange(n_eff)
tb %>% arrange(desc(Rhat))
tb %>% summarize(min = min(n_eff), max = max(n_eff))
tb %>% summarize(min = min(Rhat), max = max(Rhat))
```

# Results

Plot fixed effects.

```{r fixed_effects, fig.small=TRUE}
p_full = plot(obj, type = "beta") + 
  ggtitle(expression("Fixed Effects"~beta)) +
  xlab("log-odds of 3rd/1st trimester")
p_full
```

Extract log-odds for pSTAT1.

```{r fixed_effects_pSTAT1}
post_beta = rstan::extract(obj$fit_mcmc, pars = "beta")[[1]]
post_beta %<>% as.tibble
names(post_beta) = c("intercept",obj$protein_names)
quantile(post_beta$pSTAT1, probs = c(0.025, 0.975))
quantile(exp(post_beta$pSTAT1), probs = c(0.025, 0.975))
```

Plot random effects.

```{r random_effects, fig.small=TRUE}
plot(obj, type = "sigma_donor") + 
  ggtitle("Marker Standard Deviation"~sigma)
ggsave("sigma_glmm.pdf", width = 4, height = 3)
```

Plot posterior correlations.

```{r posterior_correlations, fig.small=TRUE}
plot(obj, type = "Cor_donor") + 
  ggtitle(expression("Marker Correlations"~Omega~"(donor)"))
ggsave("posterior_summary_cor_glmm.pdf", width = 4, height = 4)
```

# Refit Without pSTAT1

Refit model to test potentional collider confounding.

```{r refit}
protein_names = obj$protein_names[-which(obj$protein_names == "pSTAT1")]
if(file.exists(prefit_reduced)) {
  load(file = prefit_reduced)
} else {
  obj = cytoeffect::glmm(df_samples_subset, protein_names, 
                         condition = "term", group = "donor",
                         iter = 325, warmup = 200, 
                         num_chains = ncores)
  save(obj,file = prefit_reduced)
}
```

```{r plot_refit}
p_reduced = plot(obj, type = "beta") + 
  ggtitle(expression("Reduced Fixed Effects"~beta)) +
  xlab("log-odds of 3rd/1st trimester")
plot_grid(p_full, p_reduced, labels = "AUTO")
ggsave(filename = "beta_glmm.pdf", width = 8, height = 3)
```

# Session Info {.unnumbered}

```{r session_info}
sessionInfo()
```

# References {.unnumbered}
